# Simple Linear Regression

> "All models are wrong, but some are useful."\
        - **George E. P. Box**

Suppose you are the owner of the *Momma Leona's Pizza*, a restaurant chain located near several college campuses. You currently own 10 stores and have data on the size of the student population as well as quarterly sales for each. Your data is summarized in the table below.

| Restaurant | Student Population | Quarterly Sales |
|------------|--------------------|-----------------|
| 1          | 2                  | 58              |
| 2          | 6                  | 105             |
| 3          | 8                  | 88              |
| 4          | 8                  | 118             |
| 5          | 12                 | 117             |
| 6          | 16                 | 137             |
| 7          | 20                 | 157             |
| 8          | 20                 | 169             |
| 9          | 22                 | 149             |
| 10         | 26                 | 202             |

Here,

- Restaurant, $i$
- Student Population, $x_i$, in 1000s
- Quarterly Sales, $y_i$, in $1000s

```{r}
momma_leona <- data.frame(students = c(2, 6, 8, 8, 12, 16, 20, 20, 22, 26), 
                          sales = c(58, 105, 88, 118, 117, 137, 157, 169, 149, 202))
```

We have previously seen vectors and matrices for storing data as we introduced `R`. We will now introduce a **data frame** which will be the most common way that we store and interact with data in this course.

list of vectors
observations and variables
order "doesn't matter" (unlike a matrix)

```{r}
names(momma_leona)
```

```{r}
momma_leona$sales
momma_leona$students
```

```{r}
dim(momma_leona)
nrow(momma_leona)
ncol(momma_leona)
```

```{r}
str(momma_leona)
```


```{r, echo = FALSE}
write.csv(momma_leona, "data/momma_leona.csv", row.names = FALSE)
```

[data is csv here](data/momma_leona.csv)

TODO: import into R/RStudio  

```{r}
plot(momma_leona$students, momma_leona$sales)
```


```{r}
plot(momma_leona$students, momma_leona$sales,
     xlab = "Students (in 1000s)",
     ylab = "Sales (in $1000s)",
     main = "Quarterly Sales vs Student Population",
     pch  = 20,
     cex  = 2,
     col  = "dodgerblue")
```

response, outcome
predictor, explanatory

note about "independent"

How can you use this data to:

- Explain relationship
    - Significant?
    - Which *variables* (word?) are most important? (Say, a variable for public/private)
- Predict

One tool will do both, LINEAR REGRESSION

##  History

##  Model

\[
y = f(x) + \epsilon
\]

\[
y = \beta_0 + \beta_1 x + \epsilon
\]

PARAMETERS

DATA = PREDICTION + ERRORS
     = SIGNAL + NOISE
     = MODEL + ERROR

Unexplained:

- missing variables
- measurement error

##  What is the Best Line?



Goldilocks

- Underfitting (Just using y_bar)
- Just right (SLR)
- Overfitting (Connect the dots)

TODO: picture with lines. one for overall mean. one for LS?

\[
\underset{\beta_0, \beta_1}{\mathrm{argmin}} \max|y_i - (\beta_0 + \beta_1 x_i)|
\]

\[
\underset{\beta_0, \beta_1}{\mathrm{argmin}} \sum_{i = 1}^{n}|y_i - (\beta_0 + \beta_1 x_i)|
\]

\[
\underset{\beta_0, \beta_1}{\mathrm{argmin}} \sum_{i = 1}^{n}(y_i - (\beta_0 + \beta_1 x_i))^2
\]

historical, easy math

##  Least Squares Approach

Function to minimize.

\[
f(\beta_0, \beta_1) = \sum_{i = 1}^{n}(y_i - (\beta_0 + \beta_1 x_i))^2 = \sum_{i = 1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2
\]

Take derivatives.

\[
\begin{aligned}
\frac{df}{d\beta_0} &= -2 \sum_{i = 1}^{n}(y_i - \beta_0 - \beta_1 x_i) \\
\frac{df}{d\beta_1} &= -2 \sum_{i = 1}^{n}(x_i)(y_i - \beta_0 - \beta_1 x_i)
\end{aligned}
\]

Set equal to zero.

\[
\begin{aligned}
\sum_{i = 1}^{n}(y_i - \beta_0 - \beta_1 x_i) &= 0 \\
\sum_{i = 1}^{n}(x_i)(y_i - \beta_0 - \beta_1 x_i) &= 0 
\end{aligned}
\]

Rearrange, **normal equations**.

\[
\begin{aligned}
\sum_{i = 1}^{n} y_i &= n \beta_0 + \beta_1 \sum_{i = 1}^{n} x_i \\
\sum_{i = 1}^{n} x_i y_i &= \beta_0 \sum_{i = 1}^{n} x_i + \beta_1 \sum_{i = 1}^{n} x_i^2
\end{aligned}
\]

\[
\begin{aligned}
\hat{\beta}_1 &= \frac{\sum_{i = 1}^{n} x_i y_i - \frac{(\sum_{i = 1}^{n} x_i)(\sum_{i = 1}^{n} y_i)}{n}}{\sum_{i = 1}^{n} x_i^2 - \frac{(\sum_{i = 1}^{n} x_i)^2}{n}} = \frac{SXY}{SXX}\\
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1 \bar{x}
\end{aligned}
\]

\[
\begin{aligned}
SXY &= \sum_{i = 1}^{n} x_i y_i - \frac{(\sum_{i = 1}^{n} x_i)(\sum_{i = 1}^{n} y_i)}{n}  = \sum_{i = 1}^{n}(x_i - \bar{x})(y_i - \bar{y})\\
SXX &= \sum_{i = 1}^{n} x_i^2 - \frac{(\sum_{i = 1}^{n} x_i)^2}{n}  = \sum_{i = 1}^{n}(x_i - \bar{x})^2\\
SYY &= \sum_{i = 1}^{n} y_i^2 - \frac{(\sum_{i = 1}^{n} y_i)^2}{n}  = \sum_{i = 1}^{n}(y_i - \bar{y})^2
\end{aligned}
\]

\[
\hat{\beta}_1 = \frac{SXY}{SXX} = \frac{\sum_{i = 1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i = 1}^{n}(x_i - \bar{x})^2}
\]

TODO: Fitted line.
TODO: variance estimate

```{r}
momma_leona_model = lm(sales ~ students, data = momma_leona)
```

```{r}
plot(momma_leona$students, momma_leona$sales,
     xlab = "Students (in 1000s)",
     ylab = "Sales (in $1000s)",
     main = "Quarterly Sales vs Student Population",
     pch  = 20,
     cex  = 2,
     col  = "dodgerblue")
abline(momma_leona_model, lwd = 2, col = "darkorange")
```

```{r}
plot(sales ~ students, data = momma_leona,
     xlab = "Students (in 1000s)",
     ylab = "Sales (in $1000s)",
     main = "Quarterly Sales vs Student Population",
     pch  = 20,
     cex  = 2,
     col  = "dodgerblue")
abline(momma_leona_model, lwd = 2, col = "darkorange")
```

TODO: fit in R
TODO: extrapolation
TODO: interpretation

##  Decomposition of Variation

TODO: Residuals

##  Coefficient of Determination

$R^2$

\[
\sum_{i=1}^{n}(y_i - \bar{y})^2 = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2 + \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2
\]

Decomposing variation.

##  MLE Approach

\[
L(\beta_0, \beta_1, \sigma^2) = \prod_{i = 1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp{\left[-\frac{1}{2}\left(\frac{Y_i - \beta_0 - \beta_1 x_i}{\sigma}\right)^2\right]}
\]
