--- 
title: "Applied Statistics"
author: "David Dalpiaz"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
site: bookdown::bookdown_site
description: ""
url: 'https\://daviddalpiaz.github.io/appliedstats/'
github-repo: daviddalpiaz/appliedstats
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
require(knitr)
read_chunk('r_book.R')
options(scipen = 8, width = 65)
```

```{r load_packages, message=FALSE,echo=FALSE, warning=FALSE}
```

```{r misc_func, echo=FALSE}
```

# Introduction

Welcome to Applied Statistics with `R`!

## About This Book

This book was originally (and currently) designed for use with STAT 420, Methods of Applied Statistics, at the University of Illinois at Urbana-Champaign. It may certainly be used elsewhere, but you may find references to "this course" which is STAT 420.

This book is under active development. When possible, it would be best to always access the text online to be sure you are using the most up-to-date version. (Also the html version provides the most features such as changing text size, font, and colors.) If you are in need of a local copy, a pdf version is continuously maintained at <http://daviddalpiaz.github.io/appliedstats/applied_statistics.pdf>.

Since this book is under active development you may encounter errors ranging from typos to broken code to poorly explained topics. If you do, please let us know! Simply send an email and we'll make the changes ASAP. Or, if you know RMarkdown and are familiar with GitHub, [make a pull request and fix an issue yourself!](https://github.com/daviddalpiaz/appliedstats)

This text uses MathJax to render mathematical notation for the web. Occasionally but rarely a JavaScript error will prevent MathJax from rendering correctly. (And you will see the "code" instead of the expected mathematical equations.) From experience, this is almost always fixed by simply refreshing the page. You'll also notice that if you right-click any equation you can obtain the MathML Code (for copying into Microsoft Word) or the TeX command used to generate the equation.

\[
a^2 + b^2 = c^2
\]

## Conventions

`R` code will be typeset using a `monospace` font which is syntax highlighted.

```{r, eval = FALSE}
a = 3
b = 4
sqrt(a ^ 2 + b ^ 2)
```

`R` output lines, which would appear in the console will begin with `##`. They will generally not be syntax highlighted.

```{r, echo = FALSE, eval = TRUE}
a = 3
b = 4
sqrt(a ^ 2 + b ^ 2)
```

## Acknowledgements

- Alex Stepanov
- [David Unger](http://publish.illinois.edu/dunger/)
- [James Balamuta](http://www.thecoatlessprofessor.com/)