# Model Selection

> "Choose well. Your choice is brief, and yet endless."
>
> --- **Johann Wolfgang von Goethe**

After reading this chapter you will be able to:

- Understand the trade-off between goodness-of-fit and model complexity.
- Use variable selection procedures to find a good model from a set of possible models.

Last chapter we saw how correlation between predictor variables can have undesirable effects on models. We used variance inflation factors to assess the severity of the collinearity issues caused by these correlations. We also saw how fitting a smaller model, leaving out some of the correlated predictors, results in a model which no longer suffers from collinearity issues. But how should we chose this smaller model?

This chapter, we will discuss several *criteria* and *procedures* for choosing a "good" model from among a choice of many.

## Quality Criterion

So far, we have seen criteria such as $R^2$ and RMSE for assessing quality of fit. However, both of these have a fatal flaw. By increasing the size of a model, that is adding predictors, that can at worst not improve. It is impossible to add a predictor to a model and make $R^2$ or RMSE worse. (Greater, or less, respectively.) That means, if we were to use either of this to chose between models, we would *always* simply choose the larger model.

This suggests that we need a quality criteria that takes into account the size of the model, since our preference is for small models that still fit well. We are willing to sacrific a small about of goodness-of-fit for obtaining a smaller model. We will look at three criteria that do this explicitly: AIC, BIC and Adjusted $R^2$. We will also look at one, Cross-Validated RMSE, which implicitly considers the size of the model.

### Akaike Information Criterion

The first criteria we will discuss is the Akaike Information Criterion, or AIC for short. (Note that, when *Aiaike* first introduced this metric, it was simply called *An* Information Criterion. The *A* has changed meaning over the years.)

Recall, the log-likelihood of a regression model can be written as

\[
\log L(\boldsymbol{\hat{\beta}}) = -\frac{n}{2}\log(2\pi) - \frac{n}{2}\log\left(\frac{RSS}{n}\right) - \frac{n}{2},
\]

where $RSS = \sum_{i=1}^n (y_i - \hat{y}_i) ^ 2$.

Then we can define AIC as

\[
AIC = -2 \log L(\boldsymbol{\hat{\beta}}) + 2p = n + n \log(2\pi) + n \log\left(\frac{RSS}{n}\right) + 2p,
\]

which is a measure of quality of the model. The smaller the AIC, the better. To see why, let's talk about the two main components of AIC, the **likelihood** (which measures goodness-of-fit) and the **penalty** (which is a function of the size of the model).

The likelihood portion of AIC is given by

\[
-2 \log L(\boldsymbol{\hat{\beta}}) = n + n \log(2\pi) + n \log\left(\frac{RSS}{n}\right).
\]

For the sake of comparing models, the only term here that will change is $n \log\left(\frac{RSS}{n}\right)$, which is function of $RSS$. The 

\[
n + n \log(2\pi)
\]

terms will be constant across all models applied to the same data. So, when a model fits well, that is, has a low $RSS$, then this likelihood component will be small.

Similarly, we can discuss the penalty component of AIC which is,

\[
2p,
\]

where $p$ is the number of $\beta$ parameters in the model. We call this a penalty, because it is large when $p$ is large, but we are seeking to find a small AIC.

Thus, a good model, that is one with a small AIC, will have a good balance between fitting well, and using a small number of parameters. For comparing models

\[
AIC \approx n\log\left(\frac{RSS}{n}\right) + 2p
\]

is a sufficient expression, as $n + n \log(2\pi)$ is the same across all models for any particular dataset.

### Bayesian Information Criterion

The Bayesian Information Criterion, or BIC, is similar to AIC, but has a larger penalty. BIC also quantifies the trade-off between a model which fits well and the number of model parameters, however for a reasonable sample size, generally picks a smaller model than AIC. Again, for model selection use the model with the smallest BIC.

\[
BIC = -2 \log L(\boldsymbol{\hat{\beta}}) + \log(n) p = n + n\log(2\pi) + n\log\left(\frac{RSS}{n}\right) + \log(n)p.
\]

Notice that the AIC penalty was

\[
2p,
\]

whereas for BIC, the penalty is 

\[
\log(n) p.
\]

So, for any dataset where $log(n) > 2$ the BIC penalty will be larger than the AIC penalty, thus BIC will likely prefer a smaller model.  

Note that, sometimes the penalty is consider a general expression of the form

\[
k \cdot p.
\]

Then, for AIC $k = 2$, and for BIC $k = \log(n)$.

For comparing models

\[
BIC \approx n\log\left(\frac{RSS}{n}\right) + \log(n)p
\]

is again a sufficient expression, as $n + n \log(2\pi)$ is the same across all models for any particular dataset.

### Adjusted R-Squared

Recall,

\[
R^2 = 1 - \frac{SSE}{SST} = 1 - \frac{\sum_{i=1}^{n}(y_i - \hat{y}_i)^2}{\sum_{i=1}^{n}(y_i - \bar{y})^2}.
\]

We now define

\[
R_a^2 = 1 - \frac{SSE/(n-p)}{SST/(n-1)} = 1 - \left(  \frac{n-1}{n-p} \right)(1-R^2)
\]

which we call the Adjusted $R^2$.

Unlike $R^2$ which can never become smaller with added predictors, Adjusted $R^2$ effectively penalizes for additional predictors, and can decrease with added predictors. Like $R^2$, larger is still better.

### Cross-Validated RMSE

Each of the previous three metrics explicitly used $p$, the number of parameters, in their calculations. Thus, they all explicitly limit the size of models chosen when used to compare models.

We'll now briefly introduce **overfitting** and **cross-validation**.

```{r}
make_poly_data = function(sample_size = 11) {
  x = seq(0, 10)
  y = 3 + x + 4 * x ^ 2 + rnorm(n = sample_size, mean = 0, sd = 20)
  data.frame(x, y)
}
```

```{r}
set.seed(1234)
poly_data = make_poly_data()
```

Here we have generated data which follows a quadratic pattern, specifically,

\[
y = 3 + x + 4 x ^ 2 + \epsilon.
\]

We'll now fit two models to this data, one which has the correct form, quadratic, and one that is large, which includes terms up to and including an eighth degree.

```{r}
fit_quad = lm(y ~ poly(x, degree = 2), data = poly_data)
fit_big  = lm(y ~ poly(x, degree = 8), data = poly_data)
```

We then plot the data and the results of the two models.

```{r}
plot(y ~ x, data = poly_data, ylim = c(-100, 400), cex = 2, pch = 20)
xplot = seq(0, 10, by = 0.1)
lines(xplot, predict(fit_quad, newdata = data.frame(x = xplot)),
      col = "dodgerblue", lwd = 2, lty = 1)
lines(xplot, predict(fit_big, newdata = data.frame(x = xplot)),
      col = "darkorange", lwd = 2, lty = 2)
```

We can see that the solid blue curve models this data rather nicely. The dashed orange curve fits the points better, making smaller errors, however it is unlikely that it is correctly modeling the true relationship between $x$ and $y$. It is fitting the random noise. This is an example of **overfitting**.

We see that the larger model indeed has a lower RMSE.

```{r}
sqrt(mean(resid(fit_quad) ^ 2))
sqrt(mean(resid(fit_big) ^ 2))
```

To correct for this, we will introduce cross-validation. We define the leave-one-out cross-validated RMSE to be

\[
\text{CV(RMSE)}_n = \sqrt{\frac{1}{n} \sum_{i=1}^n e_{[i]}^2}.
\]

The $e_{[i]}$ are the residual for the $i$th observation, when that observation is **not** used to fit the model.

\[
{e_{[i]} = y_{i} – \hat{y}_{[i]}}
\]

That is, the fitted value is calculated as

\[
\hat{y}_{[i]} = x_i ^ \top \hat{\beta}_{[i]}
\]

where $\hat{\beta}_{[i]}$ are the estimated coefficients when the $i$th observation is removed from the dataset. 

In general, to perform this calculation, we would be required to fit the model $n$ times, once with each possible observation removed. However, for leave-one-out cross-validation and linear models, the equation can be rewritten as

\[
\text{CV(RMSE)}_n = \sqrt{\frac{1}{n}\sum_{i=1}^n [e_{i}/(1-h_{i})]^2,}
\]

where $h_i$ are the leverages and $e_i$ are the usual residuals. This is great, because now we can obtain the LOOCV RMSE by fitting only one model! In practice 5 or 10 fold cross-validation are much more popular. For example, in 5-fold cross-validation, the model is fit 5 times, each time leaving out a fifth of the data, then predicting on those values. We'll leave in-depth examination of cross-validation to a machine learning course, and simply use LOOCV here.

Let's calculate LOOCV RMSE for both models, then discuss *why* we want to do so. We first write a function which calculates the LOOCV RMSE as defined using the shortcut formula for linear models.

```{r}
calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```

Then calculate the metric for both models.

```{r}
calc_loocv_rmse(fit_quad)
calc_loocv_rmse(fit_big)
```

Now we see that the quadratic model has a much smaller LOOCV RMSE, so we would prefer this quadratic model. This is because the large model has *severely* over-fit the data. By leaving a single data point out and fitting the large model, the resulting fit is much different than the fit using all of the data. For example, let's leave out the third data point and fit both models, then plot the result.

```{r}
fit_quad_removed = lm(y ~ poly(x, degree = 2), data = poly_data[-3, ])
fit_big_removed  = lm(y ~ poly(x, degree = 8), data = poly_data[-3, ])

plot(y ~ x, data = poly_data, ylim = c(-100, 400), cex = 2, pch = 20)
xplot = seq(0, 10, by = 0.1)
lines(xplot, predict(fit_quad_removed, newdata = data.frame(x = xplot)),
      col = "dodgerblue", lwd = 2, lty = 1)
lines(xplot, predict(fit_big_removed, newdata = data.frame(x = xplot)),
      col = "darkorange", lwd = 2, lty = 2)
```

We see that on average, the solid blue line for the quadratic model has similar errors as before. It has changed very slightly. However, the dashed orange line for the large model, has a huge error at the point that was removed and is much different that the previous fit.

This is the purpose of cross-validation. By assess how the model fits points that were not used to perform the regression, we get an idea of how well the model will work for future observations. It assess how well the model works in general, not simply on the observed data.

## Selection Procedures

We've now seen a number of model quality criteria, but now we need to address which models to consider. Model selection involves both a quality criterion, plus a search procedure. 

```{r}
library(faraway)
hipcenter_mod = lm(hipcenter ~ ., data = seatpos)
coef(hipcenter_mod)
```

Let's return to the `seatpos` data from the `faraway` package. Now, let's consider only models with first order terms, thus no interactions and no polynomials. There are *eight* predictors in this model. So if we consider all possible models, ranging from using 0 predictors, to all eight predictors, there are 

\[
\sum_{k = 0}^p {{p} \choose {k}} = 2 ^ p = 2 ^ 8 = 256
\]

possible models.

If we had 10 or more predictors, we would already be considering over 100 models! For this reason, we often search through possible models in an intelligent way, bypassing some models that are unlikely to be considered good. We will consider three search procedures: backwards, forwards, and stepwise.

### Backwards Search

Backwards selection procedures start with all possible predictors in the model, then considers how deleting a single predictor will effect a chosen metric. Let's try this on the `seatpos` data. We will use the `step()` function in `R` which by default uses AIC as its metric of choice.

```{r}
hipcenter_mod_back_aic = step(hipcenter_mod, direction = "backward")
```

We start with the model `hipcenter ~ .`, which is otherwise known as `hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg`. `R` will then repeatedly attempt to delete a predictor until it stops, or reaches the model `hipcenter ~ 1`, which contains no predictors.

At each "step", `R` reports the current model, its AIC, and the possible steps with their RSS and more importantly AIC.

In this example, at the first step, the current model is `hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg` which has an AIC of `283.62`. Note that when `R` is calculating this value, it is using `extractAIC()`, which uses the expression

\[
AIC \approx n\log\left(\frac{RSS}{n}\right) + 2p,
\]

which we quickly verify.

```{r}
extractAIC(hipcenter_mod) # returns both p and AIC
n = length(resid(hipcenter_mod))
(p = length(coef(hipcenter_mod)))
n * log(mean(resid(hipcenter_mod) ^ 2)) + 2 * 9
```

Returning to the first step, `R` then gives us a row which shows the effect of deleting an of the predictors. The `-` signs at the beginning of each row indicates we are considering removing a predictor. There is also a row with `<none>` which is a row for keeping the current model. Notice that this row has the smallest RSS, as it is the largest model.

We see that every row above `<none>` has a smaller AIC than the row for `<none>` with the one at the top, `Ht`, giving the lowest AIC. Thus we remove `Ht` from the model, and continute the process.

Notice, in the second step, we start with the model `hipcenter ~ Age + Weight + HtShoes + Seated + Arm + Thigh + Leg` and the variable `Ht` is no longer considered.

We continute the process until we reach the model `hipcenter ~ Age + HtShoes + Leg`. At this step, the row for `<none>` tops the list, as removing any additional variables will not improve the AIC. This is the model which is stored in `hipcenter_mod_back_aic`.

```{r}
coef(hipcenter_mod_back_aic)
```

We could also search through the possible models in a backwards fashion using BIC. To do so, we again use the `step()` function, but now specify `k = log(n)`, where `n` stores the number of observations in the data.

```{r}
n = length(resid(hipcenter_mod))
hipcenter_mod_back_bic = step(hipcenter_mod, direction = "backward", k = log(n))
```

The procedure is exactly the same, except at each step we look to improve the BIC, which `R` is still calling AIC in the output.

The variable ``hipcenter_mod_back_bic` stores the model chosen by this procedure.

```{r}
coef(hipcenter_mod_back_bic)
```

We note that this model is *smaller*, has fewer predictors, than the model chosen by AIC, which is what we would expect.

We can use information from the `summary()` function to compare their Adjusted $R^2$ values.

```{r}
summary(hipcenter_mod_back_aic)$adj.r.squared
summary(hipcenter_mod_back_bic)$adj.r.squared
```

We can also calulate the LOOCV RMSE for both.

```{r}
calc_loocv_rmse(hipcenter_mod_back_aic)
calc_loocv_rmse(hipcenter_mod_back_bic)
```

We see that we would prefer the model chosen via BIC if using LOOCV RMSE as our metric.

### Forwards Search

```{r}
hipcenter_mod_start = lm(hipcenter ~ 1, data = seatpos)
hipcenter_mod_forw_aic = step(hipcenter_mod_start, 
                              hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg, 
                              direction = "forward")
```


```{r}
hipcenter_mod_forw_bic = step(hipcenter_mod_start, 
                              hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg, 
                              direction = "forward", k = log(n))
```

```{r}
summary(hipcenter_mod_forw_aic)$adj.r.squared
summary(hipcenter_mod_forw_bic)$adj.r.squared
```

```{r}
calc_loocv_rmse(hipcenter_mod_forw_aic)
calc_loocv_rmse(hipcenter_mod_forw_bic)
```


### Stepwise Search

- TODO: Could be reversed

```{r}
hipcenter_mod_both_aic = step(hipcenter_mod_start, 
                              hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg, 
                              direction = "both")
```

```{r}
hipcenter_mod_both_bic = step(hipcenter_mod_start, 
                              hipcenter ~ Age + Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg, 
                              direction = "both", k = log(n))
```

- TODO: neither has both Ht and HtShoes

```{r}
summary(hipcenter_mod_both_aic)$adj.r.squared
summary(hipcenter_mod_both_bic)$adj.r.squared
```

```{r}
calc_loocv_rmse(hipcenter_mod_both_aic)
calc_loocv_rmse(hipcenter_mod_both_bic)
```

### Exhaustive

- $2 ^ p$
- sum of binomial coeffs
- shortcut for methods that use RSS in formula




    


```{r}


library(leaps)
all_hipcenter_mod = summary(regsubsets(hipcenter ~ ., data = seatpos))

all_hipcenter_mod$which
which.max(all_hipcenter_mod$adjr2)

p = length(coef(hipcenter_mod))
n = length(resid(hipcenter_mod))


hipcenter_mod_aic = n * log(all_hipcenter_mod$rss / n) + 2 * (2:p) # why this equation?
which.min(hipcenter_mod_aic)
all_hipcenter_mod$which[3,]
hipcenter_mod_best_aic = lm(hipcenter ~ Age + Ht + Leg, data = seatpos)
extractAIC(hipcenter_mod_best_aic)
extractAIC(hipcenter_mod_back_aic)
extractAIC(hipcenter_mod_forw_aic)
extractAIC(hipcenter_mod_both_aic)

plot(hipcenter_mod_aic ~ I(2:p), ylab = "AIC", xlab = "p, number of parameters", 
     pch = 20, col = "dodgerblue", type = "b", cex = 2)

hipcenter_mod_bic = n * log(all_hipcenter_mod$rss / n) + log(n) * (2:p) # why this equation?
which.min(hipcenter_mod_bic)
all_hipcenter_mod$which[1,]
hipcenter_mod_best_bic = lm(hipcenter ~ Ht, data = seatpos)
extractAIC(hipcenter_mod_best_bic, k = log(n))
extractAIC(hipcenter_mod_back_bic, k = log(n))
extractAIC(hipcenter_mod_forw_bic, k = log(n))
extractAIC(hipcenter_mod_both_bic, k = log(n))

```



\[
AIC \approx n\log\left(\frac{RSS}{n}\right) + 2p.
\]

\[
BIC \approx n\log\left(\frac{RSS}{n}\right) + \log(n)p.
\]

## Higher Order Terms

`autompg` Example

- TODO: what variables are considered?
- TODO: `trace = 0`

```{r, echo = FALSE}
autompg = read.table(
  "http://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  quote = "\"",
  comment.char = "",
  stringsAsFactors = FALSE)
colnames(autompg) = c("mpg", "cyl", "disp", "hp", "wt", "acc", "year", "origin", "name")
autompg = subset(autompg, autompg$hp != "?")
autompg = subset(autompg, autompg$name != "plymouth reliant")
rownames(autompg) = paste(autompg$cyl, "cylinder", autompg$year, autompg$name)
autompg$hp = as.numeric(autompg$hp)
autompg$domestic = as.numeric(autompg$origin == 1)
autompg = autompg[autompg$cyl != 5,]
autompg = autompg[autompg$cyl != 3,]
autompg$cyl = as.factor(autompg$cyl)
autompg$domestic = as.factor(autompg$domestic)
autompg = subset(autompg, select = c("mpg", "cyl", "disp", "hp", "wt", "acc", "year", "domestic"))
```

```{r}
str(autompg)
```

```{r}
pairs(autompg)
```

- define first order, second order, etc... how they fit together

```{r}
autompg_big_mod = lm(log(mpg) ~ . ^ 2 + I(disp ^ 2) + I(disp ^ 2) + I(hp ^ 2) + I(wt ^ 2) + I(acc ^ 2), data = autompg)
autompg_mod_back_aic = step(autompg_big_mod, direction = "backward", trace = 0)

n = length(resid(autompg_big_mod))
autompg_mod_back_bic = step(autompg_big_mod, direction = "backward", k = log(n), trace = 0)

coef(autompg_mod_back_aic)
coef(autompg_mod_back_bic)


length(coef(autompg_mod_back_aic))
length(coef(autompg_mod_back_bic))

calc_loocv_rmse(autompg_mod_back_aic)
calc_loocv_rmse(autompg_mod_back_bic)

# how factors are handled

# which is better for prediction?
# which is better for explaination?
```


## Explaination versus Prediction

- TODO: its own chapter?



